Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-01 WholePlotTrajectories.tex
intro

To grasp all facets of communities response to disturbance we examined
here the taxonomic and functional trajectories in diversity and
composition \citep{Lohbeck2015, Guariguata2001}. These trajectories
would highlight the recovery of communities initial composition,
diversity and functioning and the underlying ecological processes. They
would clarify the tenants in the long term of the Intermediate
Disturbance Hypothesis debated in tropical forests, and provide
indications for future adaptive conservation strategies
\citep{Adler2007}. We monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem biomass removed. We made use of a large
functional traits database browsing major leaf, stem and seed functional
traits and species maximum height to draw the trajectories over time of
communities taxonomic and functional composition and diversity.
Specifically, we (i) questioned the coupling between taxonomic and
functional response to disturbance and identified the underlying
assembly processes, (ii) clarified the validity of the IDH in the long
term for tropical forest and (iii) questioned the completeness of
communities recovery regarding communities functional redundancy.

@@

\subsection{Inventories protocol and dataset
collection}\label{protocols}

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level,
\emph{i.e.} referring to the genus of observed or trialed botanical
names. Trajectories of communities taxonomic and functional variations
in composition after disturbance were followed in a two-dimensional NMDS
ordination space of the 30 years monitored. Two NMDS were conducted to
map either taxonomic flora inventories or communities functional
composition based on the 7 leaf, stem and life history traits (without
seed mass classes). In both cases the NMDS were performed using
abundance-based (Bray-Curtis) dissimilarity measures. Trajectories along
time were reported through the distance between successive inventories
and the reference inventories in 1989, \emph{i.e.} 2-3 years after
disturbance, when the uncertainty degree did not exceed 30\% of
undetermined trees. Taxonomic trajectories were measured through the
euclidean distance between inventories in the multivariate space of all
inventories species. Functional trajectories were measured through the
sum of euclidean distances among traits density distributions between
inventories. To compensate the intrinsinc difference among plots, the
trajectories corresponded to the euclidean differences along time with
the reference inventory in 1989. Univariate trajectories of the leaf,
stem and life-history traits were also visualized with the community
weighted means (CWM) \citep{Diaz2007, Garnier2004}. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories.

The taxonomic diversity was assessed through Richness and the Hill
number translation of Shannon and Simpson indices \citep{Hill1973}.
These three indices belong to the set of HCDT or generalized entropy,
respectively corresponding to the 0, 1 and 2 order of diversity (q),
recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the Rao index of quadratic entropy which combines
species abundance distribution and average pairwise dissimilarity based
on all functional and life traits.

The impacts of initial disturbance were first tested with the spearman
rank correlation between the extremum of taxonomic and functional
metrics reached over the 30 years and the initial \%AGB removed. Then
they were analysed through the linear correlations between Simpson and
Rao diversities and the initial \%AGB removed at 10, 20 and 30 years
after disturbance.

@@

\subsection{Communities Diversity}\label{communities-diversity}

From 1989 (2-3 years after disturbance) to 2015 (28-29 years after
disturbance), 828388 individual trees and 591 botanical species
pertaining to 223 genus and 64 botanical families were recorded. For
undisturbed plots Richness and taxonomic evenness (Shannon and Simpson
diversities) remained stable over the 30 years monitored. In disturbed
communities the taxonomic richness increased after low disturbance
intensity, reaching a maximum gain of 14 botanical genera (plot 3 from
treatment 2) while it followed unimodal trajectories after intense
disturbance, decreasing for ten years before recovering to
pre-disturbance values. In all disturbed plots the taxonomic evenness
(Shannon and Simpson diversities) increased, following unimodal
trajectories with a maximum, reached after around 20 years, positively
correlated to the disturbance treatment
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). Return towards initial evenness
values was beginning after 30 years except for two T3 plots (plots 8 and
12) which evenness still increased, suggesting similar but delayed
trajectories \ref{fig:IDHplot}.

Trajectories of communities functional diversity were examined through
the Rao diversity based on the 7 leaf, stem and life history traits (to
the exception of seed mass). The plot 7 from treatment 1 displayed a
constantly outlying diversity and was removed from the graphical
representation for better readability (see appendix for full graphs). In
undisturbed plots the functional diversity remained stable along the 30
years while in disturbed plots it followed unimodal trajectories with a
return towards initial values that strated around 20 years after
disturbance.

@@

\caption{Upper panels, Trajectories of the Simpson taxonomic diversity \textbf{(a)} and Rao functional diversity \textbf{(b)} over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \%AGB removed and Simpson \textbf{(c)} and Rao \textbf{(d)} diversities 10, 20 and 30 years after disturbance.}\label{fig:IDHplot}
\end{figure*}

@@

Both taxonomic and functional composition proved resilient and followed
similar humped-shaped trajectories starting to return towards initial
values 30 years after disturbance.

The taxonomic NMDS revealed differences in the composition of initial
communities that proved maintained throughout the recovery trajectories
after disturbance. More than commonly though, communities trajectories
after disturbance was specific to the initial composition and converged
towards its recovery
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}. The
resilience of commnities taxonomy suggested the existence of multiple
stable equilibria, corresponding here to communities initial
composition, as assumed for highly diverse and productive ecosystems
\citep{Chase2003}.

Contrastingly communities functional composition and diversity proved
more homogeneous among communities and communities functional
trajectories were similar in the functional space. Communities
functional characteristics are the most direct link between biodiversity
and ecosystem functioning \citep{Diaz2005} so their resilience meant the
recovery of ecosystem processes in the long term \citep{Guariguata2001}.

@@

The initial disturbance intensity determined the maximum functional
diversity increase, confirming the IDh within this range of disturbance.
Regarding communities communities along time, the humped-shaped
trajectories of the functional diversity following disturbance
demonstrated the positive impact of disturbance that enhanced the
diversity before a return to initial lower diversity levels. In
accordance with the IDH disturbance entailed ecological processes
different than in undisturbed communities and a temporal variability
correlated to the intensity of the disturbance. The functional diversity
increase paralleled the significant functional shifts towards
resource-acquisitive strategies (sharp increase in the SLA, leaf
thickness and bark thickness and decrease in wood density, leaf
toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. As the old-growth,
pre-disturbance survivor trees proved mirroring the initial communities
\citep{Herault2018} the functional trajectories and taxonomic turnover
rather relied upon the pool of newly recruited trees driven by the
enhanced growth and survival of previously infrequent species and
functional types. The changes in abiotic environment and competitive
pressure favored pioneers which outcompeted other species in non
limiting light resources but were excluded in mature forests by
long-lived, more resistant and shade tolerant species.
\textgreater{}\textgreater{} Not clear Disturbance trajectories
therefore relied upon the environmental niches made available and filled
by species from a restricted functional range that became dominant and
structured the functional characteristics of the community
\citep{Grime1998}.

In contrast the trajectories of taxonomic richness and evenness were
decoupled from the disturbance intensity, questioning the consistency of
the IDH to predict communities taxonomic diversity, as already observed
in the Guiana Shield \citep{Baraloto2012a} and in Bornean tropical
forests \citep{Cannon1998}.

In the most disturbed plots the taxonomic richness did not exceed this
of the initial communities but the taxonomic evenness significantly
increased, revealing the loss of most infrequent species. Dominant
species fastly recovered and restored the communities functoinal
characteristics that mostly rely on dominant species, according to the
``vegetation quantity effect'' \citep{Grime1998}, while the recovery of
more infrequent species was hampered. \textgreater{}\textgreater{} Not
clear The intermediate disturbance hypothesis was suitabe to predict
communities functional diversity and pool of dominant species
\citep{Molino2001}, but it poorly represented the whole taxonomic
structure that was hampered by the slow recovery of rare species
\citep{Hubbell2001, Chave2004}.

\subsection{The functional redundancy, key of communities
resilience}\label{the-functional-redundancy-key-of-communities-resilience}

The recovery of the most infrequent species was hampered by the
competition among species, in accordance with the lottery recruitment
\citep{Busing2002} and entailed long term recovery cycles
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The progressive increase of the functional redundancy within the initial
functional space, that had first sharply decreased, explains the lottery
recruitment that first allow an easy taxonomic recovery and then a
progressive slow down. functional redundancy restricted to the
functional space of the initial community clearly followed. After
disturbance communities differently occupied the functional space,
following the observed shifts in traits trajectories and functional
composition, and the redundancy in the initial functional space
decreased according to the disturbance intensity. After 30 years the
recovery of the initial redundancy structure was consistent but remained
unachieved.

\begin{quote}
\begin{quote}
unclear The persistent alteration of functional redundancy meant a lower
resilience of pre-disturbance communities and an higher resilience
disturbance-specific communities (corresponding to lower richness and
higher dominance of pioneer species) so chances to observe
self-maintained compositional changes in favor of disturbance resistant
species, lianas or epiphytes were increased after disturbance
\citep{Haddad2008, Burslem2000, Martin2013}.
\end{quote}
\end{quote}

\begin{quote}
\begin{quote}
unclear (cornerstone?) This decades-long recovery specifically increased
the risks of loosing cornerstone specie, with unexpected ecological
consequences \citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}.
Indeed the flora-faune relationship and the functional traits studied
are not exhaustive and key function might be lost. specifically for the
fauna-flora relationships.
\end{quote}
\end{quote}

@@

Our study defined the decoupled functional and taxonomic trajectories of
tropical rainforests after disturbance demonstrating a rapid functional
recovery but a slower and more variable taxonomic one. Consistently with
the IDH, functional trajectories were driven after disturbance by
enrichment of the system with pioneers and light-demanding species.

\begin{quote}
\begin{quote}
unclear The following functional shifts entailed a re-organization of
communities functional redundancy that proved long to recover and
explained the more variable and longer taxonomic trajectories as shade
tolerant species of old-growth forests faced functional redundancy and
limiting similarity processes.
\end{quote}
\end{quote}

The resilience of tropical forests then proved long in the face of quite
intense disturbance and did not preclude the settling of a persistent
disturbance-specific community \citep{Gourlet-Fleury2005}. As the
trajectories highlighted the recruitment processes proved central for
communities response to disturbance and closer focus on demographical
drivers of communities response would clarify the fate of the future
forests.

\begin{quote}
\begin{quote}
unclear The disturbance range however stayed within the spectrum of
selective logging, with a forest cover remaining all along the
experiment, and the response mechanisms would probably be much different
after harder disturbance.
\end{quote}
\end{quote}

>>>>

To grasp all facets of communities response to disturbance we examined
here the taxonomic and functional trajectories in terms of richness,
evennes and composition \citep{Lohbeck2015, Guariguata2001}. These
trajectories would highlight the recovery of communities initial
characteristics and the underlying ecological processes. They would
clarify the tenants in the long term of the Intermediate Disturbance
Hypothesis debated in tropical forests, and provide indications for
future adaptive conservation strategies \citep{Adler2007}. We monitored
over 30 years the response of 75 ha of neotropical forest plots set up
on a gradient of disturbance intensity, from 10 to 60\% of ecosystem
biomass removed. We made use of a large functional traits database
browsing major leaf, stem and seed functional traits and species maximum
height to draw the trajectories over time of communities taxonomic and
functional composition,richness and evenness. Specifically, we (i)
questioned the coupling between taxonomic and functional response to
disturbance and identified the underlying assembly processes, (ii)
clarified the validity of the IDH in the long term for tropical forest
and (iii) questioned the completeness of communities recovery regarding
communities functional redundancy.

@@

\subsection{Inventories protocol and dataset
collection}\label{inventories-protocol-and-dataset-collection}

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level,
\emph{i.e.} referring to the genus of observed or trialed botanical
names. Trajectories of communities taxonomic and functional variations
in composition after disturbance were followed in a two-dimensional NMDS
ordination space of the 30 years monitored. Two NMDS were conducted to
map either taxonomic flora inventories or communities functional
composition based on the 7 leaf, stem and life history traits (without
seed mass classes). In both cases the NMDS were performed using
abundance-based (Bray-Curtis) dissimilarity measures. Trajectories along
time were reported through the distance between successive inventories
and the reference inventories in 1989, \emph{i.e.} 2-3 years after
disturbance, when the uncertainty degree did not exceed 30\% of
undetermined trees. for the taxonomix composition trajectories were
measured by the euclidean distance between inventories in the
multivariate space of all inventories species. For the functoinal
composition trajectories were measured by the sum for all traits of the
euclidean distances between the traits density distributions of
successive inventories. To compensate the intrinsinc difference among
plots, the trajectories corresponded to the euclidean differences along
time with the reference inventory in 1989. Univariate trajectories of
the leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \citep{Diaz2007, Garnier2004}. Species
seed mass corresponded to 5 classes of increasing mass, seed mass
trajectories were therefore reported as the proportion of each class in
the inventories.

The taxonomic diversity was reported through species richness and the
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These two indices belong to the set of HCDT or
generalized entropy, respectively corresponding to the 0 and 2 order of
diversity (q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the Rao index of quadratic entropy which combines
species abundance distribution and average pairwise dissimilarity based
on all functional and life traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extremum of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB removed. The were
besides analysed through the linear correlations between Simpson and Rao
diversities and the initial \%AGB removed at 10, 20 and 30 years after
disturbance.

@@

\subsection{Communities richness and
evenness}\label{communities-richness-and-evenness}

From 1989 (2-3 years after disturbance) to 2015 (28-29 years after
disturbance), 828388 individual trees and 591 botanical species
pertaining to 223 genus and 64 botanical families were recorded. For
undisturbed plots taxonomic Richness and Evenness (Simpson diversity)
remained stable over the 30 years monitored. In disturbed communities
after low disturbance intensity the taxonomic richness increased,
reaching a maximum gain of 14 botanical genera (plot 3 from treatment
2). After intense disturbance the taxonomic richness followed unimodal
trajectories, decreasing for ten years after disturbance before
recovering to pre-disturbance values. In all disturbed plots the
taxonomic evenness (Simpson diversity) followed unimodal trajectories
with first an increase until a maximum reached after around 20 years
that was positively correlated to the disturbance treatment
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). The return towards initial values
of communities evenness was beginning after 30 years except for two T3
plots (plots 8 and 12) which evenness still increased, suggesting
similar but delayed trajectories \ref{fig:DivTaxo}.

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/DivTaxo-1}

\caption{Trajectories of the difference to the 1989 inventories (5 years after disturbance) over 30 years after disturbance of plots communities \textbf{(a)} Richness, \textbf{(b)} Shannon and \textbf{(c)} Simpson diversities. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed a constantly outlying functoinal
evenness and was removed from the graphical representation for better
readability (see appendix for full graphs). In undisturbed plots the
functional evenness remained stable along the 30 years while in
disturbed plots it followed unimodal trajectories with a return towards
initial values starting around 20 years after disturbance.

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/FunDiv-1}

\caption{Trajectories of the Rao functional diversity over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:FunDiv}
\end{figure}

@@

\caption{Upper panels, Trajectories of the taxonomic evenness (Simpson diversity) \textbf{(a)} and Rao functional diversity \textbf{(b)} over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \%AGB removed and Simpson \textbf{(c)} and Rao \textbf{(d)} diversities 10, 20 and 30 years after disturbance.}\label{fig:IDHplot1}
\end{figure*}\begin{figure*}

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/IDHplot-2}

\caption{Upper panels, Trajectories of the taxonomic evenness (Simpson diversity) \textbf{(a)} and Rao functional diversity \textbf{(b)} over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \%AGB removed and Simpson \textbf{(c)} and Rao \textbf{(d)} diversities 10, 20 and 30 years after disturbance.}\label{fig:IDHplot2}
\end{figure*}

@@

Communities taxonomic and functional composition and functional evenness
proved resilient, following similar humped-shaped trajectories starting
to return towards pre-disturbance composition and functioning after 30
years.

The taxonomic differences among communities, marked before disturbance
by the distinct starting points of the NMDS, were maintained throughout
communities recovery trajectories. More than commonly thought
post-disturbance trajectories depended on communities initial
composition, that partly determined the pool of rectuitable species and
that constrained the converge towards the recovery of the initial
composition
\citep{Hubbell1999, Molino2001, Anderson2007, Baraloto2012a}. The
resilience of communities taxonomy suggested the existence of multiple
stable equilibria, here the pre-disturbance communities, as assumed for
highly diverse and productive ecosystems \citep{Chase2003}.

Communities functional composition followed similar trajectories that
were not so distinct in the functional space. The trajectories of
functional evenness as well were similar among communities Communities
functional characteristics are the most direct link between biodiversity
and ecosystem functioning and their resilience here meant the recovery
of ecosystem processes and services ona cycle of several decades
\citep{Diaz2005, Guariguata2001}.

@@

The IDH consistently predicted communities functional response to
disturbance and the functional trajectories in time highlighted the
underlying ecological mechanisms, but the theory was diproved regarding
communities taxonomic richness and evenness.

With the studied range of intensity, disturbance enhanced communities
functional evenness but did not have clear impact on taxonomic richness
or evenness. The IDH proved a good predictor of the disturbance impact
on communities functioning but not on their taxonomic richness or
evenness.

The humped-shaped trajectories of communities along time demonstrated
the enhanced functional evennes and changes in functional composition
following disturbance before a return towards initial, lower values. In
accordance with the IDH, disturbance entailed significant temporal
variability with the emergence of ecological processes different from
those rueling undisturbed forest. The increase in functional evenness
went with significant functional shifts towards resource-acquisitive
strategies (sharp increase in the SLA, leaf thickness and bark thickness
and decrease in wood density, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Because the pre-disturbance
survivor trees mirror the initial communities \citep{Herault2018}, the
functional changes relied upon recruited trees and the enhanced growth
and survival of previously infrequent species and functional types.
Pioneers outcompeted other species in the post-disturbance environment
of non limiting resources and alleviated competition. Before disturbance
and over communities recovery time though, these species would be
excluded by long-lived, more resistant and shade tolerant species.

In contrast the taxonomic richness and evenness were decoupled from the
disturbance intensity, questioning the consistency of the IDH as already
observed in the Guiana Shield \citep{Baraloto2012a} and in Bornean
tropical forests \citep{Cannon1998}. The recovery of communities
functional characteristics, that mostly rely on dominant species
according to the ``vegetation quantity effect'' \citep{Grime1998},
suggested a similar recovery of the dominant species \citep{Molino2001}.
Still, the taxonomic trajectories were blurried by the hampered recovery
of infrequent species that entailed after disturbance a decrease in the
taxonomic richness and an increase of the evenness
\citep{Hubbell2001, Chave2004}. The disturbance therefore porrly
predicted commuities taxonomic richnessa and evennes that rely upon
different ecological processes.

\subsection{The functional redundancy, key of the taxonomic
resilience}\label{the-functional-redundancy-key-of-the-taxonomic-resilience}

The recovery of the most infrequent species followed the lottery
recruitment rules \citep{Busing2002} and was hampered by the increasing
competition among species whih entailed long term recovery cycles
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The rapid decrease of the functional redundancy within the initial
functional space, correlated with the disturbance intensity, illustrated
the space and resources made available after disturbance. Following the
lottery recruitment, communities taxonomic richness increased firsthand
along the enhanced recruitment rate but decreased thereafter, following
the emergence of exclusive competition. After 30 years the recovery of
the initial redundancy structure was underway but remained unachieved.
This alteration of the functional redundancy meant a lower resilience of
pre-disturbance communities and higher chances to see the persistence of
disturbance-specific communities, with less species and more pioneers
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the hampered
recovery of rare species increased the risks to loose cornerstone
specie, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species mighthave
unique functions i nthe ecosystem or be a key for some fauna.

@@

Our study revealed communities cyclic recovery after disturbance
entailing the resilience of their functioning and taxonomic composition
with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for 20 years after
disturbance through the enrichment of the communities with pioneers and
light-demanding species,in accordance with the IDH. The IDH, though,
poorly predicted the disturbance impact on communities taxonomic
richness and evenness that were blurried by the emergence of competitive
exclusion along time. The resilience of tropical forests proved
consistent although spread over several decades. Still, the disturbance
impact on communities redundancy cautioned against the risks of
infrequent species loss and the persistence of disturbance-specific
communities \citep{Gourlet-Fleury2005}. As the trajectories highlighted
the recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.
